---
title: "4. Case Study: An Inflamed Fibroblast State in Spatial Colitis Data"
author:
    - name: Anthony Christidis
      affiliation:
        - &core_affiliation Core for Computational Biomedicine, Harvard Medical School
    - name: Andrew Ghazi
      affiliation:
        - *core_affiliation
    - name: Smriti Chawla
      affiliation:
        - *core_affiliation
    - name: Nitesh Turaga
      affiliation:
        - *core_affiliation
    - name: Ludwig Geistlinger
      affiliation:
        - *core_affiliation
    - name: Robert Gentleman
      affiliation:
        - *core_affiliation
package: scDiagnostics
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{4. Case Study: An Inflamed Fibroblast State in Spatial Colitis Data}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include = FALSE}
knitr::knit_hooks$set(pngquant = knitr::hook_pngquant)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "ragg_png",
  dpi = 72,
  fig.retina = 2,
  fig.align = "center",
  out.width = "100%",
  pngquant = "--speed=1 --quality=1-5"
)
```

# Purpose

[Vignettes 2](https://ccb-hms.github.io/scDiagnostics/articles/ZeiselBenchmarking.html)
and [3](https://ccb-hms.github.io/scDiagnostics/articles/COVIDCaseStudy.html)
worked with dissociated scRNA-seq data. `scDiagnostics` operates on
`r BiocStyle::Biocpkg("SingleCellExperiment")` objects and does not
assume any particular assay - the same diagnostics apply directly to a
`r BiocStyle::Biocpkg("SpatialExperiment")` object from imaging-based
spatial transcriptomics, without modification, and the same holds for
`r BiocStyle::Biocpkg("SpatialFeatureExperiment")` objects (which extend
`SpatialExperiment` with explicit cell/tissue geometries), since both
inherit the same core `SingleCellExperiment` interface that
`scDiagnostics` relies on. This vignette repeats the project-detect-
characterize workflow on MERFISH spatial data from a mouse model of
DSS-induced colitis (Cadinu et al. 2024): healthy colon tissue (Day 0,
reference) versus tissue at peak inflammation (Day 9, query).

```{r, message=FALSE}
library(scDiagnostics)
library(SingleCellExperiment)
library(SpatialExperiment)
```

# The data

`merfish_reference_data` and `merfish_query_data` are downsampled
subsets on the same 943-gene targeted MERFISH panel, restricted to 5
shared cell types; see `?merfish_reference_data` for details. At Day 9,
some fibroblasts further split into an inflammation-associated
"Inflamed Fibroblast" state (`tier2_merged`) that does not exist at Day
0; a merged `cell_type_merged` column collapses this back into a shared
"Fibroblast" label so the two timepoints can be compared directly.

```{r, message=FALSE}
data("merfish_reference_data")
data("merfish_query_data")

class(merfish_reference_data)
table(merfish_reference_data$cell_type_merged)
table(merfish_query_data$tier2_merged[merfish_query_data$cell_type_merged == "Fibroblast"])
```

# Step 1: project

```{r, fig.height=6, fig.width=10}
plotCellTypePCA(
    query_data = merfish_query_data,
    reference_data = merfish_reference_data,
    cell_types = unique(merfish_reference_data$cell_type_merged),
    query_cell_type_col = "cell_type_merged",
    ref_cell_type_col = "cell_type_merged",
    pc_subset = 1:3)
```

# Step 2: detect

```{r, fig.height=5, fig.width=10}
anomaly_output <- detectAnomaly(
    reference_data = merfish_reference_data,
    query_data = merfish_query_data,
    ref_cell_type_col = "cell_type_merged",
    query_cell_type_col = "cell_type_merged",
    cell_types = "Fibroblast",
    pc_subset = 1:5,
    n_tree = 500)

is_anomalous <- anomaly_output[["Fibroblast"]]$query_anomaly
mean(is_anomalous)
```

```{r, fig.height=5, fig.width=10}
plot(anomaly_output, cell_type = "Fibroblast", pc_subset = 1:3, data_type = "query")
```

Because `merfish_query_data` retains the original fine-grained
`tier2_merged` label, we can check how the flagged cells relate to the
ground-truth "Inflamed Fibroblast" state:

```{r, message=FALSE}
fibro_query <- merfish_query_data[, merfish_query_data$cell_type_merged == "Fibroblast"]
tapply(is_anomalous, fibro_query$tier2_merged, mean)
```

In this downsampled dataset, cells with the ground-truth "Inflamed
Fibroblast" label are flagged as anomalous roughly twice as often as
plain "Fibroblast" cells - a real enrichment, though far from a clean
separation, consistent with inflammation being a graded rather than
binary state at the single-cell level.

# Step 3: characterize

To characterize what distinguishes the flagged cells, `calculateGeneShifts()`
can run its own internal anomaly detection (`detect_anomalies = TRUE`,
`anomaly_comparison = TRUE`) on the full 943-gene panel - the same
panel `detectAnomaly()` used above - while restricting the actual
statistical comparison to a small extracellular matrix (ECM) gene panel
via `genes_to_analyze`. This keeps detection and characterization
consistent without needing to manually subset cells first:

```{r, message=FALSE}
ecm_signature <- c("Col1a2", "Timp2", "Col6a1", "Sparc", "Dpt")

gene_shifts <- calculateGeneShifts(
    query_data = merfish_query_data,
    reference_data = merfish_reference_data,
    query_cell_type_col = "cell_type_merged",
    ref_cell_type_col = "cell_type_merged",
    cell_types = "Fibroblast",
    pc_subset = 1:5,
    genes_to_analyze = ecm_signature,
    detect_anomalies = TRUE,
    anomaly_comparison = TRUE)

gene_shifts$PC1[order(gene_shifts$PC1$p_adjusted), ]
```

`plot()` shows this as a heatmap of per-gene z-scores, one column per
cell (reference and query, annotated by anomaly status), rather than
collapsing each group to a single averaged column - which matters here,
since it reveals more than the group averages alone would:

```{r, fig.height=6, fig.width=8}
plot(gene_shifts, cell_type = "Fibroblast", pc_subset = 1:5,
    plot_type = "heatmap", plot_by = "p_adjusted", n_genes = 5,
    show_anomalies = TRUE)
```

All five ECM genes are significantly lower in the anomalous query
fibroblasts than in the reference, but the heatmap shows this isn't a
single uniform effect. `Col1a2` and `Sparc` show a visible extra drop
concentrated specifically in the anomalous (rightmost, red-annotated)
cells, beyond what's already true of the query more broadly. `Timp2`,
`Col6a1`, and `Dpt`, on the other hand, are already substantially
reduced across essentially *all* query fibroblasts - anomalous or not -
so for those three genes, `detectAnomaly()` isn't isolating a
distinctly-shifted subgroup so much as reflecting a shift already
present dataset-wide. This split held up across several PC-subset
choices we checked, so it looks like a real property of this dataset
rather than a detection parameter to tune away: some genes distinguish
the specific cells flagged as anomalous, and others distinguish the
query condition as a whole - both are useful, but they're different
claims.

The barplot below focuses on `Col1a2` and `Sparc` specifically, since
they're the clearest example of the anomaly-specific pattern - a
"Query Non-Anomaly vs Ref" bar close to zero alongside a much larger
"Query Anomaly vs Ref" bar, showing the shift really is concentrated in
the flagged cells for these two genes (unlike `Timp2`/`Col6a1`/`Dpt`
above, where the non-anomaly bar would already be nearly as large as
the anomaly bar):

```{r, fig.height=5, fig.width=8}
gene_shifts_focused <- calculateGeneShifts(
    query_data = merfish_query_data,
    reference_data = merfish_reference_data,
    query_cell_type_col = "cell_type_merged",
    ref_cell_type_col = "cell_type_merged",
    cell_types = "Fibroblast",
    pc_subset = 1:5,
    genes_to_analyze = c("Col1a2", "Sparc"),
    detect_anomalies = TRUE,
    anomaly_comparison = TRUE)

plot(gene_shifts_focused, cell_type = "Fibroblast", pc_subset = 1:5,
    plot_type = "barplot", plot_by = "p_adjusted", n_genes = 2,
    show_anomalies = TRUE, pseudo_bulk = TRUE)
```

As with the COVID-19 case study, this is a specific finding about this
cell population in this dataset; the broader claim that this workflow
generalizes across data modalities is best supported by comparing this
result to the scRNA-seq case study in vignette 3, not by either result
alone.

------------------------------------------------------------------------

# R Session Info

```{r SessionInfo, echo=FALSE, message=FALSE, warning=FALSE, comment=NA}
options(width = 80)
sessionInfo()
```
